---
title: "Data Science CYO Project"
author: "Robert Lewis"
date: "June 13, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project Introduction
Working for a Universities which offers OSHA Trainer Training of which we offer online and live courses. Our online safety courses are very short, but numerous. Our live classroon classes cover all saftey aspects within different indutrusties such as Construction, General Industry, Maritime and Disaters. I an effort to increase live class room enrollments, The purpose of this project (CYO) is to create an algorithm for recommending classes for users based on other programs they have completed and the scores they received.  I will be using a subset of modified data to protect the actual users infomation. I will then use the recommenderlab for making my recommendation and using different models then attempt to identify the best model with the lowest error acuracy.


```{r Install Packages and libraries, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(useful)) install.packages("useful", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
if(!require(recommenderlab)) install.packages("recommenderlab", repos = "http://cran.us.r-project.org")
library(dplyr)
library(stringr)
library(tidyverse)
library(caret)
library(data.table)
library(useful)
library(lubridate)
library(recommenderlab)

